\subsection{Statistical precision and power}
\label{sec:results:precision}

A negative result is only as informative as the precision behind it, so we
report what our data can and cannot rule out rather than the trial counts
alone. Throughout we attach exact one-sided $95\%$ Clopper--Pearson upper
bounds to per-condition and pooled failure rates, and we report the minimum
per-condition accuracy drop we could have detected at $80\%$ power, following
the eval-uncertainty practice of \citet{miller2024errorbars}. The precision is
deliberately and substantially \emph{uneven} across the design, and the
strength of any individual claim should be read off the relevant bound below,
not off the aggregate accuracy.

\paragraph{Aggregate precision.} Pooling the registered grid across the four
models (\texttt{gpt-5.5}, \texttt{gpt-5.4}, \texttt{gpt-5.4-mini},
\texttt{claude-sonnet-4-6}; easy and hard probe sets) gives $7{,}330$
present-needle trials with $48$ failures, for an overall present-needle
accuracy of $0.9935$ and an exact one-sided $95\%$ Clopper--Pearson upper bound
on the failure rate of $0.87\%$. This pooled bound is tight, but it aggregates
over heterogeneous cells and over the single non-null cell
(Table~\ref{tab:b2}); it should not be read as a per-condition guarantee, since
length-driven degradation is a per-condition effect that pooling can mask.

\paragraph{Per-condition precision is coarse for most cells.} Resolving the
design to its finest grain---a condition is a model $\times$ probe set $\times$
probe $\times$ context length cell, present-needle only ($120$ conditions)---the
per-condition sample size ranges from $n=2$ to $n=432$, with a median of $24$.
The corresponding exact one-sided $95\%$ Clopper--Pearson upper bounds on the
per-condition failure rate have a median of $0.149$, a best (largest cell,
$n=432$) of $0.0085$, and a worst (smallest cell, $n=2$) of $0.842$. Expressed
as detectable effects, the minimum per-condition accuracy drop we could detect
at $80\%$ power has a median of $0.065$, a best of $0.0037$ ($n=432$), and a
worst of $0.553$ ($n=2$). For the smallest cells the data exclude only large
per-condition collapses; a drop of, say, ten points within such a cell would
frequently go undetected. We therefore do not interpret a flat $1.000$ in a
small cell as a tightly bounded per-condition result.

\paragraph{One arm is tightly powered; the others are breadth.} The
per-condition precision is concentrated almost entirely in one arm. Pooled
present-needle sample sizes are $6{,}030$ for \texttt{gpt-5.4-mini}, $720$ for
\texttt{gpt-5.5}, $600$ for \texttt{gpt-5.4}, and $170$ for
\texttt{claude-sonnet-4-6}. The depth of the design lives on
\texttt{gpt-5.4-mini}: its largest cells reach $n=432$, where the failure-rate
upper bound is $0.0085$ and the minimum detectable per-condition drop is
$0.0037$. On that model, and only on that model, we can state that no
length-driven per-condition degradation larger than roughly $0.4\%$ is
consistent with the data on our controlled, synthetic, mechanically-scored
probes up to 150k tokens. The frontier-tier arms (\texttt{gpt-5.5},
\texttt{gpt-5.4}, \texttt{claude-sonnet-4-6}) are \emph{breadth} confirmation,
not tightly-powered per-condition nulls: their cells are small (as small as
$n=2$), so they can exclude only large per-condition drops. Their role is to
show that the depth-arm pattern is not idiosyncratic to a single model or
provider, not to bound small effects on each frontier model individually.

\paragraph{What this supports.} Read together, the evidence supports a bounded
statement: on these controlled, difficulty-invariant, mechanically-scored
probes, for the four models tested and up to 150k tokens, we observe no
measurable length-driven degradation, with the strength of that statement set
by the bounds above---tight on the depth arm (\texttt{gpt-5.4-mini}, drops
$>\!\sim\!0.4\%$ per condition excluded) and coarse but broad across the
frontier-tier arms (only large per-condition drops excluded). The lone
exception, HARD~B2 aggregation on \texttt{gpt-5.4-mini}
(Table~\ref{tab:b2}), is itself measured in the best-powered cells, which is
why a dip of order ten points there is detectable while comparable drops in
small frontier-tier cells would not be. Two boundaries remain even at this
precision. First, the controlled \texttt{end} arm equalizes the needle--probe
distance but does not fully disentangle position, distance, and filler
composition, so a small residual confound cannot be excluded. Second, this is
a synthetic harness whose inputs may be cleaner than real multi-turn agentic
workflows; these per-condition bounds describe raw input volume on controlled
tasks and do not estimate volume's contribution in deployed agents, and they
may age as model versions change.
